Variational Monte Carlo on Bosonic Systems

### 2018-04-04 project_1/article/article.tex
Restore old crap

\subsection{Non-interacting harmonic oscillators}
    We start by looking at a simple system of non-interacting harmonic
    oscillators. That is, where $a = 0$ and $\beta = 1$, i.e. there is no hard-shell
    interaction between particles and the particles are confined in a spherical trap.
    We thus get the trial  wavefunction
    \begin{align}
        \Psi_T(\vf{r})
        = \Phi_T(\vf{r})
        = \prod_{i = 1}^N \exp\bigl[
            -\alpha |\vf{r}_i|^2
        \bigr],
    \end{align}
    where $|\vf{r}_i| = r_i$. As $a = 0$ the interaction term,
    $w(\vf{r}_i, \vf{r}_j)$, vanishes and the Hamiltonian is given by
    (in the spherical case)
    \begin{align}
        H &= \sum_{i = 1}^N h(\vf{r}_i)
        = \sum_{i = 1}^N \Biggl(
            -\frac{\hbar^2}{2m}\nabla_i^2
            + \half m \omega^2 r_i^2
        \Biggr).
    \end{align}
    To find the drift force and the local energy we have to compute the gradient
    and the Laplacian of the trial wavefunction. The gradient is given by
    \begin{align}
        \nabla_k\Psi_T(\vf{r})
        &= -2\alpha \vf{r}_k\Psi_T(\vf{r}),
    \end{align}
    whereas the Laplacian yields
    \begin{align}
        \nabla^2_k\Psi_T(\vf{r})
        &= \big(-2d\alpha + 4\alpha^2 r_k^2\bigr)\Psi_T(\vf{r}),
    \end{align}
    where $d$ is the dimensionality of the problem determined by
    $\vf{r}_k \in \mathbb{R}^d$. We can thus use the gradient to find an
    expression for the drift force for particle $k$.
    \begin{align}
        \vf{F}_k(\vf{r})
        &= -4\alpha\vf{r}_k.
    \end{align}
    Using the Laplacian we can compute the kinetic term in the expression for the
    local energy. We get
    \begin{align}
        E_L(\vf{r})
        &=
        \sum_{i = 1}^N
        \Biggl(
            -\frac{\hbar^2}{2m}
            \bigl[
                -2d\alpha + 4\alpha^2 r_i^2
            \bigr]
            + \half m\omega^2 r_i^2
        \Biggr).
    \end{align}
    In natural units, with $\hbar = c = m = 1$, this reduces to
    \begin{align}
        E_L(\vf{r})
        &=
        \alpha dN
        + \biggl(
            \half\omega^2
            - 2\alpha^2
        \biggr)
        \sum_{i = 1}^N r_i^2.
        \label{eq:closed_form_natural_units_local_energy}
    \end{align}
    It is worth noting that for $\alpha = \half\omega$ ($\alpha$ is required to
    be positive) we will find a stable value which turns out to be the exact
    energy minimum. This happens as the sum over all particles
    disappears.

    \subsubsection{Exact variational energy}
        As the system is non-interacting and consisting of Gaussians we can
        find an expression for the exact energy as a function of the
        variational parameter $\alpha$, i.e.,
        \begin{align}
            E(\alpha)
            &=
            \frac{\bra{\Psi_T}H\ket{\Psi_T}}{\bra{\Psi_T}\ket{\Psi_T}}.
        \end{align}

        % TODO: Add the computation of this expression
        The final result for the energy is
        \begin{align}
            E(\alpha)
            &=
            \left(
                \frac{\hbar^2 \alpha}{2m}
                + \frac{m\omega^2}{8\alpha}
            \right)dN.
            \label{eq:exact_energy}
        \end{align}
        By minimizing this expression, i.e., finding the derivative of the
        energy with respect to $\alpha$ and equating this to zero, yields the
        expected minimum of variational energy to be
        \begin{align}
            \dod[]{E(\alpha)}{\alpha} = 0
            \implies
            \alpha_0 = \frac{m\omega}{2\hbar},
        \end{align}

        The drift force for particle $k$ in this system becomes

        \begin{align}
            E(\alpha_0)
            &=
            \frac{\omega dN}{2}.
        \end{align}

        The Laplacian of particle $k$ divided by the trial wavefunction is
        \begin{widetext}
            \begin{align}
                \frac{\nabla_k^2\Psi_T(\vf{r})}{\Psi_T(\vf{r})}
                &=
                \frac{\nabla_k^2\phi_k}{\phi_k}
                + 2\frac{\nabla_k\phi_k}{\phi_k}
                \sum_{m \neq k}^N
                \frac{\Delta\vf{r}_{km}}{r_{km}}
                \dpd[]{u_{km}}{r_{km}}
                + \sum_{m\neq k}^N
                \para{
                    \frac{d - 1}{r_{km}}\dpd[]{u_{km}}{r_{km}}
                    + \dpd[2]{u_{km}}{r_{km}}
                }
                +
                \para{
                    \sum_{m \neq k}^N
                    \frac{\Delta\vf{r}_{km}}{r_{km}}
                    \dpd[]{u_{km}}{r_{km}}
                }^2,
                \label{eq:laplacian_full_system}
            \end{align}
        \end{widetext}
        where $\Delta \vf{r}_{km} = \vf{r}_k - \vf{r}_m$ and $d$ is the
        dimensionality. Evaluating the gradient and Laplacian of the single
        particle functions yields
        \begin{gather}
            \frac{\nabla_k\phi_k}{\phi_k}
            + \sum_{m \neq k}^N
            \nabla_k u_{km}
        		\Biggr)
       		 \Psi_T(\vf{r}).
        		\label{eq:gradient_full_wavefunction}
    		\end{gather}

    The Laplcian of the trial wavefunction is found by finding the divergence of
    \autoref{eq:gradient_full_wavefunction}.
    \begin{align}
        \nabla_k^2\Psi_T(\vf{r})
        &=
        \Biggl(
            \nabla_k\Biggl[
                \frac{\nabla_k\phi_k}{\phi_k}
            \Biggr]
            +
            \sum_{m \neq k}^N \nabla_k^2 u_{km}
        \Biggr)\Psi_T(\vf{r})
        \\
        &\qquad
        +
        \Biggl(
            \frac{\nabla_k\phi_k}{\phi_k}
            + \sum_{m \neq k}^N
            \nabla_k u_{km}
        \Biggr)^2
        \Psi_T(\vf{r}),
    \end{align}
    where the squared term came from taking the gradient of the trial
    wavefunction.  To further simplify we divide by the trial
    wavefunction. This yields
    \begin{align}
        \frac{\nabla_k^2\Psi_T(\vf{r})}{\Psi_T(\vf{r})}
        &=
        \frac{\nabla_k^2\phi_k}{\phi_k}
        + 2\frac{\nabla_k\phi_k}{\phi_k}
        \sum_{m \neq k}\nabla_k u_{km}
        \nonumber \\
        &\qquad
        + \sum_{m\neq k}^N\nabla_k^2 u_{km}
        + \Biggl(
            \sum_{m \neq k}^N\nabla_k u_{km}
        \Biggr)^2.
    \end{align}
    From here, the next step is to find the gradient and the Laplacian of
    the single particle functions, $\phi_k$, and the interaction
    functions $u_{km}$. For the single particle functions we use
    Cartesian coordinates when finding the derivatives while for
    the interaction functions we use spherical coordinates and do a
    variable substitution. Beginning with the gradient of the single
    particle functions we get
    \begin{align}
        \nabla_k\phi_k
        &=
        \nabla_k\exp\bigl[
            -\alpha(x_k^2 + y_k^2 + \beta z_k^2)
        \bigr] \\
        &=
        -2\alpha
        (x_k\vf{e}_i + y_k\vf{e}_j + \beta z_k\vf{e}_k)
        \phi_k.
    \end{align}
    Note that the subscripts on the unit vectors $\vf{e}_i$ are
    \emph{not} the same as the subscripts used for its components. The
    Laplacian yields
    \begin{align}
        \nabla_k^2\phi_k
        &=
        \Bigl[
            -2\alpha
            \big(
                d - 1 + \beta
            \bigr)
            \nonumber \\
            &\qquad
            + 4\alpha^2
            \bigl(
                x_k^2 + y_k^2 + \beta^2z_k^2
            \bigr)
        \Bigr]
        \phi_k,
    \end{align}
    with $d$ as the dimensionality of the problem.
    In order to derive the interaction functions we have to do a
    variable substitution using $r_{km} = |\vf{r}_k - \vf{r}_m|$. We can
    then rewrite the $\nabla_k$-operator as
    \begin{align}
        \nabla_k
        &=
        \nabla_k
        \dpd[]{r_{km}}{r_{km}}
        =
        \nabla_k r_{km} \dpd[]{}{r_{km}}
        \\
        &=
        \frac{\vf{r}_k - \vf{r}_m}{r_{km}}\dpd[]{}{r_{km}}.
    \end{align}
    Applying this version of the $\nabla_k$-operator to $u_{km}$ yields
    \begin{align}
        \nabla_k u_{km}
        &=
        \frac{\vf{r}_k - \vf{r}_m}{r_{km}}
        \dpd[]{u_{km}}{r_{km}}.
    \end{align}

@@

        The local energy is thus
        \begin{align}
            E_L(\vf{r})
            &=
            \sum_{k = 1}^N
            \brac{
                -\frac{\hbar^2}{2m}
                \frac{\nabla_k^2\Psi_T(\vf{r})}{\Psi_T(\vf{r})}
                + v(\vf{r}_k)
            },
        \end{align}
        where the Laplcian is given by \autoref{eq:laplacian_full_system}.  We
        will avoid writing out the full expression for the local energy for this
        system due to the size of the equation.

        Looking at the one-body part of the Hamiltonian we get
        \begin{align}
            H
            &=
            \sum_{k = 1}^N
            \brac{
                -\frac{\hbar^2}{2m}\nabla_k^2
                + \half m\brak{
                    \omega^2(x_k^2 + y_k^2)
                    + \omega_z^2 z_k^2
                }}
                \\
                &=
                \frac{\hbar\omega}{2}
                \sum_{k = 1}^N
                \brac{
                    -{\nabla_k'}^2
                    +
                    \brak{
                        {x_k'}^2 + {y_k'}^2 + \lambda^2{z_k'}^2
                    }
             	    },
                \label{eq:scaled_hamiltonian}
            \end{align}
            where we have introduced the dimensionless frequency $\lambda =
            \omega_z/\omega$. The single particle functions also get a scaling
            factor
            \begin{align}
                \phi_k
                &=
                \exp\brak{
                    -\frac{\alpha\hbar}{m\omega}
                    ({x_k'}^2 + {y_k'}^2 + \beta^2{z_k'}^2)
                }
                \equiv
                \phi(\vf{r}_k').
                \label{eq:scaled_spf}
            \end{align}
            The correlation wavefunction and the interaction potential remain the
            same, but using the scaled hard core radius $a/a_{\text{ho}} =
            0.0043$.

>>>>

    \subsection{Interacting hard sphere bosons}
        Moving to the full elliptical system for $\beta \neq 1$ and setting $a
        \neq 0$ we can write the trial wavefunction as
        \begin{align}
            \Psi_T(\vf{r})
            &=
            \Phi_T(\vf{r})
            J(\vf{r}),
        \end{align}
        where $\Phi_T(\vf{r})$ is the same Slater permanent as in
        \autoref{eq:initial_trial_wavefunction} and $J(\vf{r})$ is the
        \emph{Jastrow factor} given by
        \begin{align}
            J(\vf{r})
            &=
            \exp\para{
                \sum_{j < l}^N u(r_{jl})
            },
        \end{align}
        where
        \begin{align}
            u(r_{jk})
            = \ln\bigl[f(a, \vf{r}_j, \vf{r}_k)\bigr]
            \equiv u_{jk}
        \end{align}
        The drift force for particle $k$ in this system becomes
        \begin{align}
            \vf{F}_k(\vf{r})
            &=
            2\Biggl(
                \frac{\nabla_k\phi_k}{\phi_k}
                + \sum_{m \neq k}^N
                \nabla_k u_{km}
            \Biggr).
        \end{align}

        The Laplacian of particle $k$ divided by the trial wavefunction is
        \begin{widetext}
            \begin{align}
                \frac{\nabla_k^2\Psi_T(\vf{r})}{\Psi_T(\vf{r})}
                &=
                \frac{\nabla_k^2\phi_k}{\phi_k}
                + 2\frac{\nabla_k\phi_k}{\phi_k}
                \sum_{m \neq k}^N
                \frac{\Delta\vf{r}_{km}}{r_{km}}
                \dpd[]{u_{km}}{r_{km}}
                + \sum_{m\neq k}^N
                \para{
                    \frac{d - 1}{r_{km}}\dpd[]{u_{km}}{r_{km}}
                    + \dpd[2]{u_{km}}{r_{km}}
                }
                +
                \para{
                    \sum_{m \neq k}^N
                    \frac{\Delta\vf{r}_{km}}{r_{km}}
                    \dpd[]{u_{km}}{r_{km}}
                }^2,
                \label{eq:laplacian_full_system}
            \end{align}
        \end{widetext}
        where $\Delta \vf{r}_{km} = \vf{r}_k - \vf{r}_m$ and $d$ is the
        dimensionality. Evaluating the gradient and Laplacian of the single
        particle functions yields
        \begin{gather}
            \frac{\nabla_k\phi_k}{\phi_k}
            =
            -2\alpha
            (x_k\vf{e}_i + y_k\vf{e}_j + \beta z_k\vf{e}_k),
            \\
            \frac{\nabla_k^2\phi_k}{\phi_k}
            =
            -2\alpha
            \big(
                d - 1 + \beta
            \bigr)
            + 4\alpha^2
            \bigl(
                x_k^2 + y_k^2 + \beta^2z_k^2
            \bigr).
        \end{gather}
        Repeating this for the correlation functions yields
        \begin{align}
            \dpd[]{u_{km}}{r_{km}}
            &=
            \frac{a}{r_{km}(r_{km} - a)},
            \\
            \dpd[2]{u_{km}}{r_{km}}
            &= \frac{a^2 - 2ar_{km}}{r_{km}^2(r_{km} - a)^2}.
        \end{align}
        The local energy is thus
        \begin{align}
            E_L(\vf{r})
            &=
            \sum_{k = 1}^N
            \brac{
                -\frac{\hbar^2}{2m}
                \frac{\nabla_k^2\Psi_T(\vf{r})}{\Psi_T(\vf{r})}
                + v(\vf{r}_k)
            },
        \end{align}
        where the Laplcian is given by \autoref{eq:laplacian_full_system}.  We
        will avoid writing out the full expression for the local energy for this
        system due to the size of the equation.

@@

            Looking at the one-body part of the Hamiltonian we get
            \begin{align}
                H
                &=
                \sum_{k = 1}^N
                \brac{
                    -\frac{\hbar^2}{2m}\nabla_k^2
                    + \half m\brak{
                        \omega^2(x_k^2 + y_k^2)
                        + \omega_z^2 z_k^2
                    }
                }
                \\
                &=
                \frac{\hbar\omega}{2}
                \sum_{k = 1}^N
                \brac{
                    -{\nabla_k'}^2
                    +
                    \brak{
                        {x_k'}^2 + {y_k'}^2 + \lambda^2{z_k'}^2
                    }
                },
                \label{eq:scaled_hamiltonian}
            \end{align}
            where we have introduced the dimensionless frequency $\lambda =
            \omega_z/\omega$. The single particle functions also get a scaling
            factor
            \begin{align}
                \phi_k
                &=
                \exp\brak{
                    -\frac{\alpha\hbar}{m\omega}
                    ({x_k'}^2 + {y_k'}^2 + \beta^2{z_k'}^2)
                }
                \equiv
                \phi(\vf{r}_k').
                \label{eq:scaled_spf}
            \end{align}
            The correlation wavefunction and the interaction potential remain the
            same, but using the scaled hard core radius $a/a_{\text{ho}} =
            0.0043$.
